Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-14 manuscript.tex
Update on Overleaf.

It is worth highlighting that myelinization becomes increasingly important in larger brains where conductance delays are substantial and brainwave rhythms are critical; synchrony errors can lead to neuropsychiatric and neurological dysfunctions~\citep{fields_new_2015}, such as Parkinson's disease, epilepsy or multiple sclerosis~\citep{madadi_asl_dendritic_2018}. Additionally, a recent study suggests that demyelination of the optic nerve could be an underlying factor in glaucoma~\citep{xue_2021}. ~\citet{duncan_neuron-oligodendrocyte_2021} states that "the loss of myelin and oligodendrocytes fundamentally alters the neuron, [which are then] susceptible to energetic failure [and] subsequent degeneration".

@@

However, one question remains: how do oligodendrocytes detect neuronal activity and regulate the myelinization accordingly? To answer this, we must first study the myelinization process. The oligodendrocyte precursor cells (OPCs) first proliferate in the white matter \textit{via} a self-repulsive process, thus allowing for an evenly spaced network, and identify target axons. Most OPCs then differentiate into oligodendrocytes and immediately initiate myelinization, with no further migration~\citep{simons_oligodendrocytes_2016}. Not much is known about how oligodendrocytes select the axons to myelinate, but it seems that myelination only takes place on large enough axons and is strongly regulated by several factors~\citep{kuhn_2019}, such as $Ca^{2+}$ activity~\citep{baraban_2018} or the neuregulin 1 growth factor~\citep{nave_2006}. The important role of myelination on delay learning and biological behaviors, as highlighted in the previous paragraph, suggests that the identification of target neurons as well as the myelin production is also regulated by neuronal activity. Indeed, \citet{cullen_2021} experimentally demonstrated that learning and associated neuronal activity modify the Ranvier nodes' length and the periaxonal space width in the adult mouse brain. They also confirmed that the delay correlates with the level of skill acquisition. \citet{gibson_2014} suggests that neuronal activity does not solely promote adaptive myelination in the mammalian brain, but also OPC differentiation and oligodendrogenesis. Some further studies show that oligodendrocytes may detect neuronal activity thanks to growth factors or neurotransmitters released through ion channels or \textit{via} exocytosis, but does not require any axo-glial synaptic communication~\citep{fields_2015}.

A potential scenario for selective myelination on electrically active axons using non synaptic junctions between an axon and an oligodendrocyte is: the axon releases glutamate in the extracellular environment by vesicle fusion, which activate the oligodendrocyte's NMDA and metabotropic glutamate receptors. This triggers the axo-glial signalling complex, involving the phosphorylation of the SRC family kinase FYN followed by the translation of heterogeneous nuclear ribonucleoprotein A2 into local myelin basic proteins~\citep{fields_2015}.

% myelination for axonal delay, but what about dendritic delay ? 
The previous paragraphs present the biological mechanisms behind axonal delay, regulated by myelination. However, dendritic delay seems to play an equally important role in the precise timing within a sequence of spikes: dendritic delay is involved in the performance, structure and function of the nervous system, the modulation of spatio-temporal properties of pre- and post-synaptic activity patterns and the functional limitations of sensory feedback control efficiency~\citep{MadadiAsl_2018}. Its role has been specifically identified in the compensation of input asynchrony in the mammalian auditory brainstem~\citep{spencer_2018}. \citet{branco_dendritic_2010} demonstrates the dendritic sensitivity to a sequence of synaptic activation in cortical pyramid neurons, encoded by "both local dendritic calcium signals and somatic depolarization, leading to sequence-selective spike output". The dendritic mechanism described can identify patterns delivered to a single dendrite or randomly distributed across the dendritic tree, and relies on the dendritic calcium influx moderation by NMDA receptors.

>>>>

It is worth highlighting that myelinization becomes increasingly important in larger brains where conductance delays are substantial and brainwave rhythms are critical; synchrony errors can lead to neuropsychiatric and neurological dysfunctions~\citep{fields_new_2015}, such as Parkinson's disease, epilepsy or multiple sclerosis~\citep{madadi_asl_dendritic_2018}. Additionally, a recent study suggests that demyelination of the optic nerve could be an underlying factor in glaucoma~\citep{xue_demyelination_2021}. ~\citet{duncan_neuron-oligodendrocyte_2021} states that "the loss of myelin and oligodendrocytes fundamentally alters the neuron, [which are then] susceptible to energetic failure [and] subsequent degeneration".

@@

However, one question remains: how do oligodendrocytes detect neuronal activity and regulate the myelinization accordingly? To answer this, we must first study the myelinization process. The oligodendrocyte precursor cells (OPCs) first proliferate in the white matter \textit{via} a self-repulsive process, thus allowing for an evenly spaced network, and identify target axons. Most OPCs then differentiate into oligodendrocytes and immediately initiate myelinization, with no further migration~\citep{simons_oligodendrocytes_2016}. Not much is known about how oligodendrocytes select the axons to myelinate, but it seems that myelination only takes place on large enough axons and is strongly regulated by several factors~\citep{kuhn_oligodendrocytes_2019}, such as $Ca^{2+}$ activity~\citep{baraban_ca2_2018} or the neuregulin 1 growth factor~\citep{nave_axonal_2006}. The important role of myelination on delay learning and biological behaviors, as highlighted in the previous paragraph, suggests that the identification of target neurons as well as the myelin production is also regulated by neuronal activity. Indeed, \citet{cullen_periaxonal_2021} experimentally demonstrated that learning and associated neuronal activity modify the Ranvier nodes' length and the periaxonal space width in the adult mouse brain. They also confirmed that the delay correlates with the level of skill acquisition. \citet{gibson_neuronal_2014} suggests that neuronal activity does not solely promote adaptive myelination in the mammalian brain, but also OPC differentiation and oligodendrogenesis. Some further studies show that oligodendrocytes may detect neuronal activity thanks to growth factors or neurotransmitters released through ion channels or \textit{via} exocytosis, but does not require any axo-glial synaptic communication~\citep{fields_new_2015}.

A potential scenario for selective myelination on electrically active axons using non synaptic junctions between an axon and an oligodendrocyte is: the axon releases glutamate in the extracellular environment by vesicle fusion, which activate the oligodendrocyte's NMDA and metabotropic glutamate receptors. This triggers the axo-glial signalling complex, involving the phosphorylation of the SRC family kinase FYN followed by the translation of heterogeneous nuclear ribonucleoprotein A2 into local myelin basic proteins~\citep{fields_new_2015}.

% myelination for axonal delay, but what about dendritic delay ? 
The previous paragraphs present the biological mechanisms behind axonal delay, regulated by myelination. However, dendritic delay seems to play an equally important role in the precise timing within a sequence of spikes: dendritic delay is involved in the performance, structure and function of the nervous system, the modulation of spatio-temporal properties of pre- and post-synaptic activity patterns and the functional limitations of sensory feedback control efficiency~\citep{madadi_asl_dendritic_2018}. Its role has been specifically identified in the compensation of input asynchrony in the mammalian auditory brainstem~\citep{spencer_compensation_2018}. \citet{branco_dendritic_2010} demonstrates the dendritic sensitivity to a sequence of synaptic activation in cortical pyramid neurons, encoded by "both local dendritic calcium signals and somatic depolarization, leading to sequence-selective spike output". The dendritic mechanism described can identify patterns delivered to a single dendrite or randomly distributed across the dendritic tree, and relies on the dendritic calcium influx moderation by NMDA receptors.
